Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia

### 2024-10-23 analysis/supplementary-materials/SOM_datadictionaries.qmd
Added variables to better correspond to the other SI documents

tibble::tribble(
  ~Variable, ~"Allowed values", ~Description, ~Reference,
  "Type", "-", "Chert type ID.", "-",
  "Sub-type", "-", "Chert subtype ID.", "-",
  "Color", "-", "Color according to the Munsell Soil Color Chart.", "-",
  "Fabric", "Homogeneous, Heterogeneous", "Distribution of color, luster, translucency, feel and pattern in the sample. **Heterogeneous**: presence of different characteristics within one single sample.", "@luedtke_archaeologists_1992",
  "Luster", "Shiny, Medium, Dull", "The character of light reflected by minerals. **Shiny**: very intense luster; **Medium**: some luster but it is not very intense; **Dull**: no luster.", "@crandell_macroscopic_2005-1 and @luedtke_archaeologists_1992",
  "Translucency - qualitative", "Highly translucent, Translucent, Sub-translucent, Opaque", "Degree to which the light can penetrate a material when held 5cm from a 40w lamp. **Highly translucent**: transparent or near transparent; **Translucent**: silhouettes can be seen through the sample’s thin part and/or light passes through the thick parts; **Sub-translucent**: light only passes through the thin parts; **Opaque**: no or almost no light passes through.", "@bressyCaracterisationGestionSilex2002 and @crandell_macroscopic_2005-1",
  "Translucency - quantitative", "-", "Measurement in mm of the thickness where light begins to pass through in translucent to sub-translucent samples.", "-",
  "Feel", "Smooth, Semi-smooth, Rough", "A fingernail is dragged across the sample’s surface. **Smooth**: nothing can be felt; **Semi-smooth**: slight rough feel; **Rough**: if the sample is distinctly rough.", "@bressyCaracterisationGestionSilex2002 and @crandell_macroscopic_2005-1",
  "Pattern", "Shaded, Spotted, Splotched, Broad mottling, Marbled mottling, Speckling, Flecks, Banded, Streaked, Laminated, Finely laminated", "Type of pattern present on the sample. **Shaded**: gradient color or texture change; **Spotted**: circle patterns less than 30% of the surface area; **Splotched**: irregular shapes less than 30% of the surface area; **Broad mottling**: large, irregular blotching, covering more than 30% of the surface; **Marbled mottling**: large relatively round shapes, covering more than 30% of the surface; **Speckling**: small dots, well distributed over the surface; **Flecks**: small dots, grouped together; **Banded**: regular lines thicker than 1 cm; **Streaked**: less regular and wider banding; **Laminated**: lines with thickness inferior to 1 cm; **Finely laminated**: series of lines with thickness inferior to 1 mm.", "@crandell_macroscopic_2005-1",
  "Cortex type", "Outcrop, Rounded, Pebble, Indeterminate", "**Outcrop**: uneven and low level of alteration to the original cortical surface; **Rounded**: homogeneous and rounded cortex, with possible slight alteration to the original cortical surface; **Pebble**: Smooth and highly weathered surface, with high degree of alteration to the cortical surface.", "-",
  "Cortex maximum thickness", "-", "Measurement in mm of the thickest cortex area.", "-",
  "Cortex minimum thickness", "-", "Measurement in mm of the thinnest cortex area.", "-",
  "Cortex transition", "Sharp, Gradual", "**Sharp**: transition between cortex and chert is sharp, with an obvious change in colors and textures; **Gradual**: cortex and chert show a gradual transition with gradient colors and textures", "-",
  "Allochems", "-", "Macroscopically visible allochems (excluding bioclasts).", "-",
  "Bioclasts", "-", "Macroscopically visible bioclasts.", "-",
  "Fracture", "Conchoidal, Uneven", "Presence or absence of a conchoidal fracture.", "-",
  "Quality", "Good, Acceptable, Low", "Combination of: 1) homogeneity, understood as the lack of fractures, flaws and cracks; 2) fine granularity and smooth feel; 3) lack of inclusions that due to their nature or distribution impact the knapping.", "@luedtke_chert_1978 and @brandl_genesis_2013"
) |>
  knitr::kable()

>>>>

  "Type", "-", "Chert type ID.", "-",
  "Sub-type", "-", "Chert subtype ID.", "-",
  "Color", "-", "Color according to the Munsell Soil Color Chart.", "-",
  "Fabric", "Homogeneous, Heterogeneous", "Distribution of color, luster, translucency, feel and pattern in the sample. **Heterogeneous**: presence of different characteristics within one single sample.", "@luedtke_archaeologists_1992",
  "Luster", "Shiny, Medium, Dull", "The character of light reflected by minerals. **Shiny**: very intense luster; **Medium**: some luster but it is not very intense; **Dull**: no luster.", "@crandell_macroscopic_2005-1 and @luedtke_archaeologists_1992",
  "Translucency - qualitative", "Highly translucent, Translucent, Sub-translucent, Opaque", "Degree to which the light can penetrate a material when held 5cm from a 40w lamp. **Highly translucent**: transparent or near transparent; **Translucent**: silhouettes can be seen through the sample’s thin part and/or light passes through the thick parts; **Sub-translucent**: light only passes through the thin parts; **Opaque**: no or almost no light passes through.", "@bressyCaracterisationGestionSilex2002 and @crandell_macroscopic_2005-1",
  "Translucency - quantitative", "-", "Measurement in mm of the thickness where light begins to pass through in translucent to sub-translucent samples.", "-",
  "Feel", "Smooth, Semi-smooth, Rough", "A fingernail is dragged across the sample’s surface. **Smooth**: nothing can be felt; **Semi-smooth**: slight rough feel; **Rough**: if the sample is distinctly rough.", "@bressyCaracterisationGestionSilex2002 and @crandell_macroscopic_2005-1",
  "Pattern", "Shaded, Spotted, Splotched, Broad mottling, Marbled mottling, Speckling, Flecks, Banded, Streaked, Laminated, Finely laminated", "Type of pattern present on the sample. **Shaded**: gradient color or texture change; **Spotted**: circle patterns less than 30% of the surface area; **Splotched**: irregular shapes less than 30% of the surface area; **Broad mottling**: large, irregular blotching, covering more than 30% of the surface; **Marbled mottling**: large relatively round shapes, covering more than 30% of the surface; **Speckling**: small dots, well distributed over the surface; **Flecks**: small dots, grouped together; **Banded**: regular lines thicker than 1 cm; **Streaked**: less regular and wider banding; **Laminated**: lines with thickness inferior to 1 cm; **Finely laminated**: series of lines with thickness inferior to 1 mm.", "@crandell_macroscopic_2005-1",
  "Cortex type", "Outcrop, Rounded, Pebble, Indeterminate", "**Outcrop**: uneven and low level of alteration to the original cortical surface; **Rounded**: homogeneous and rounded cortex, with possible slight alteration to the original cortical surface; **Pebble**: Smooth and highly weathered surface, with high degree of alteration to the cortical surface.", "-",
  "Cortex maximum thickness", "-", "Measurement in mm of the thickest cortex area.", "-",
  "Cortex minimum thickness", "-", "Measurement in mm of the thinnest cortex area.", "-",
  "Cortex transition", "Sharp, Gradual", "**Sharp**: transition between cortex and chert is sharp, with an obvious change in colors and textures; **Gradual**: cortex and chert show a gradual transition with gradient colors and textures", "-",
  "Morphology of the nodules", "Irregular or rounded nodules, tabular, bedded, lenses", "Inferred morphology of the chert based on cortical samples, unknapped chunks or preserved geological features such as encasing rock attached.", "@trewinSEDIMENTARYROCKSChert2005",
  "Allochems", "-", "Macroscopically visible allochems (excluding bioclasts).", "-",
  "Bioclasts", "-", "Macroscopically visible bioclasts.", "-",
  "Fracture", "Conchoidal, Uneven", "Presence or absence of a conchoidal fracture.", "-",
  "Quality", "Good, Acceptable, Low", "Combination of: 1) homogeneity, understood as the lack of fractures, flaws and cracks; 2) fine granularity and smooth feel; 3) lack of inclusions that due to their nature or distribution impact the knapping.", "@luedtke_chert_1978 and @brandl_genesis_2013"
